Ore Polynomials in Sage

### 2013-06-10 paper/main.tex
paper: references added, section on ore polynomials

Algorithms for Ore algebras provide a systematic answer to this
question~\cite{..,..}.  Invented in the early 20th century~\cite{..} with the
objective of providing a unified theory for various kinds of linear operators,
they have been used for many years in computer algebra systems, for example in
the Maple packages gfun~\cite{..} and mgfun~\cite{..}, in the Mathematica
packages by Mallinger~\cite{..} and Koutschan~\cite{..,..}, and
elsewhere~\cite{..}.

The purpose of this paper is to introduce an implementation of a collection of
algorithms related to Ore algebras for the computer algebra system
Sage~\cite{..}. It is addressed to first-time users who are already familiar
with Sage, and with the theory of Ore algebras and its use for doing symbolic
computation related to special functions. Readers unfamiliar with Sage are referred
to~\cite{..}, and readers unfamiliar with Ore algebras may wish to consult the
recent tutorial~\cite{..} and the references given there for an introduction to
the subject.

@@

More details on the construction of Ore algebras are given in the following section.
Also Ore algebras with several generators are already supported. However, so far we offer
no functionality beyond addition and multiplication for these. More functionality 
is available for algebras with one generator. We describe some of it in 
Section~\ref{sec:3}. Even more functionality has been implemented for special 
algebras like the ones created above, which are most frequently used in practice. 
See Section~\ref{sec:4} for an overview. Finally, the package also comes with a 
fine tuned guessing engine, whose main features we describe in Section~\ref{sec:5}.

@@

As another example, here is how to define an algebra of differential operators 
whose base ring is a differential field $K=\set Q(x,y,z)$ where $y$ represents
$\exp(x)$ and $z$ represents $\log(x)$:

@@

arithmetic, lclm, gcrd, action, coercion and conversion, pretty printing, 
coefficient extraction, closure properties, associate solutions.

\section{Special Methods for Standard Algebras}\label{sec:4}

special methods available for univ algebras with univ algebra as ground ring

\subsection{Differential Operators}

D and theta

\subsection{Recurrence Operators}

delta and S

\subsection{The $q$-Case}

J and Q

\section{Guessing}\label{sec:5}

[[CITE RUBEY.]]

Guessing is one of the most popular features of software packages like gfun.  It
is a technique for detecting patterns in sequences of numbers, which is often
used in combinatorics. The basic idea is simple. If we are given a finite array
of numbers, though of as the first terms of an infinite sequence, we may want to
know whether this sequence satisfies a recurrence. The algorithm behind a
guessing engine searches for small equations matching the given data.
Generically, no such equations exist, so if some are found, it is fair to
``guess'' that they are in fact valid equations for the whole infinite sequence.

@@

For the most typical situations, there are two important hyperbolas. One
describes the region in the $(r,d)$-plane consisting of all points for which
there exists an operator of order~$r$ and degree~$d$ truly satisfied by the
sequence in question. (See~\cite{..} for an explanation why the boundary of this
region is usually a hyperbola.) The second describes the region of all points
$(r,d)$ for which an operator of order~$r$ and degree~$d$ can be detected when
$N$ terms are provided as input. This region is determined by the requirement
$(r+1)(d+2)<N$.

>>>>

Algorithms for Ore algebras provide a systematic answer to this
question~\cite{bronstein96,chyzak98}.  Invented in the first half of the 20th
century~\cite{ore33} with the objective of providing a unified theory for
various kinds of linear operators, they have been used for many years in
computer algebra systems, for example in the Maple packages
OreTools~\cite{abramov03}, gfun~\cite{salvy94} or mgfun~\cite{chyzak98a}, in the
Mathematica packages by Mallinger~\cite{mallinger96} and
Koutschan~\cite{koutschan10c,koutschan09}.

The purpose of this paper is to introduce an implementation of a collection of
algorithms related to Ore algebras for the computer algebra system
Sage~\cite{sage}. It is addressed to first-time users who are already familiar
with Sage, and with the theory of Ore algebras and its use for doing symbolic
computation related to special functions. Readers unfamiliar with Sage are referred
to~\cite{sage}, and readers unfamiliar with Ore algebras may wish to consult the
recent tutorial~\cite{kauers13} and the references given there for an introduction to
the subject.

@@

More details on the construction of Ore algebras are given in the following section.
Also Ore algebras with several generators are already supported. However, so far we offer
no functionality beyond addition and multiplication for these. More functionality 
is available for algebras with one generator. We describe some of it in 
Section~\ref{sec:3}. Even more functionality has been implemented for special 
algebras like the ones created above, which are most frequently used in practice. 
Section~\ref{sec:4} describes some of these features.

@@

As another example, here is how to define an algebra of differential operators 
whose base ring is a differential field $K=\set Q(x,y,z)$ where $y$ represents
$\exp(x)$ and $z$ represents~$\log(x)$:

@@

Ore polynomials are elements of Ore algebras, i.e., Sage objects whose parent is an Ore algebra
object as described in the previous section. They can be constructed by addition and multiplication
from generators and elements of the base ring.

\begin{sageexample}
  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
  sage: (5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)
\end{sageexample}

Alternatively, an Ore polynomial can be constructed from any piece of data that is also accepted
by the constructor of the associated commutative algebra. The associated commutative algebra of
an Ore algebra is the commutative polynomial ring with the same base ring as the Ore algebra 
and with generators that are named like the generators of the Ore algebra. In particular, it is
possible to create an Ore polynomial from the corresponding commutative polynomial, from a 
coefficient list, or even from a string representation.

\begin{sageexample}
  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
  sage: Ac = A.associated_commutative_algebra(); Ac
  sage: A(Ac.random_element())
  sage: A([5*x,7*x-3,3*x+1])
  sage: A("(5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)")
\end{sageexample}

Ore polynomials can also be created from Ore polynomials that belong to other algebras, provided
that such a conversion is meaningful.

\begin{sageexample}
  sage: R.<x> = ZZ['x']; A.<Dx> = OreAlgebra(R)
  sage: L = (5*x^2+3*x-7)*Dx^2 + (3*x^2+8*x-1)*Dx + (9*x^2-3*x+8)
  sage: L.parent()
  sage: B = OreAlgebra(QQ['x'], 'Dx')
  sage: L = B(L)
  sage: L.parent()
\end{sageexample}

In accordance with the default Sage coercion model, such conversions take place automatically (if possible) 
when operators from different algebras are added or multiplied. Note that the result of such an operation
need not belong to either of the parents of the operands but may instead have a suitable ``common extension'' 
as parent.

\begin{sageexample}
  sage: A = OreAlgebra(ZZ['t']['x'], 'Dx'); 
  sage: B = OreAlgebra(QQ['x'].fraction_field(), 'Dx')
  sage: L = A.random_element() + B.random_element()
  sage: L.parent()
\end{sageexample}

\section{Selected Methods}\label{sec:4}

\subsection{gcrd and lclm?}

-- options for selecting algorithms / remainder sequence

-- cite max's paper

-- available for all (univariate) algebras

\subsection{closure properties?}

-- table mapping function names to arithmetic

-- example for how to prove some identity with the methods provided.

\subsection{solutions?}

-- polynomial, rational, generalized series solutions

-- available for differential and shift algebras

\subsection{action?}

-- should be mentioned somewhere, although there is not much to say about it.

\subsection{to list?}

-- options for handling singularities

-- rectangular splitting

-- available for shift algebras.

\subsection{Guessing}

Guessing is a technique for detecting patterns in sequences of numbers, which is
often used in combinatorics. It is one of the most popular features of packages
like gfun, and there are even some special purpose packages dedicated to this
technique~\cite{kauers09a,hebisch11}.  The basic idea is simple. If we are given
a finite array of numbers, though of as the first terms of an infinite sequence,
we may want to know whether this sequence satisfies a recurrence. The algorithm
behind a guessing engine searches for small equations matching the given data.
Generically, no such equations exist, so if some are found, it is fair to
``guess'' that they are in fact valid equations for the whole infinite sequence.

@@

For the most typical situations, there are two important hyperbolas. One
describes the region in the $(r,d)$-plane consisting of all points for which
there exists an operator of order~$r$ and degree~$d$ truly satisfied by the
sequence in question. (See~\cite{jaroschek13a} for an explanation why the
boundary of this region is usually a hyperbola.) The second describes the region
of all points $(r,d)$ for which an operator of order~$r$ and degree~$d$ can be
detected when $N$ terms are provided as input. This region is determined by the
requirement $(r+1)(d+2)<N$.
